---
title: "CMP330 Data Analytics - Assessment 2 (Pre-Assignment)"
author: "Md Jahirul Islam -> B00999362"
output:
  html_notebook: default
  pdf_document: default
---

# Objective

The task is to predict a car's acceptability (`unacc`, `acc`, `good`, 
`vgood`) from six attributes describing the car. I'll go through the usual 
EDA steps first, then compare five ML algorithms, and finally try to squeeze 
out some extra accuracy through feature selection and hyperparameter tuning.

---

# Part 1: Exploring and Preparing the Data

## 1.1 Reading the file in

The target and all predictors are categorical, so everything gets converted 
to a factor right after loading - otherwise R will treat some columns as 
plain character strings and a lot of the functions below won't behave 
properly.

```{r}
eval_data <- read.csv("car.csv")
eval_data <- data.frame(lapply(eval_data, factor))

head(eval_data)
str(eval_data)
```

## 1.2 Basic counts

Just wanted to get a feel for the data before doing anything fancy - how 
balanced are the classes, and does anything jump out from a couple of 
cross-tabs.

```{r}
# class balance - turns out "unacc" dominates, which makes sense since most
# combinations of features probably don't produce a "great" car
table(eval_data$acceptability)

table(eval_data$buying, eval_data$acceptability)

# curious whether low-safety cars are basically never rated highly
table(eval_data$safety, eval_data$acceptability)

# and whether 2-seaters get penalised regardless of everything else
table(eval_data$persons, eval_data$acceptability)
```

## 1.3 Some plots

I went with a few different chart types here instead of just bar charts, to 
get a couple of different angles on the same data.

```{r}
library(ggplot2)

eval_data$acceptability <- factor(eval_data$acceptability,
                                   levels = c("unacc", "acc", "good", "vgood"))

# plain bar chart of the target - good starting point
ggplot(eval_data, aes(x = acceptability)) +
  geom_bar(fill = "#4A6FA5") +
  labs(title = "Cars by Acceptability Rating", x = "Acceptability", y = "Number of cars") +
  theme_minimal()

# stacked proportion chart instead of dodge, to see the *relative* mix of
# classes within each buying price band rather than raw counts
ggplot(eval_data, aes(x = buying, fill = acceptability)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#B0413E", "#DDA448", "#3C6E71", "#284B63")) +
  labs(title = "Proportion of Acceptability Classes within each Buying Price",
       x = "Buying Price", y = "Proportion") +
  theme_minimal()

# faceted view - safety broken out into its own small panels, one per class
ggplot(eval_data, aes(x = safety, fill = safety)) +
  geom_bar() +
  facet_wrap(~ acceptability) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Safety Levels Split Out by Acceptability Class", x = "Safety", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 1.4 Mutual information (instead of correlation)

None of these variables are numeric so normal correlation won't work. 
Mutual information is the categorical equivalent - roughly, how much 
knowing one variable narrows down what the other one could be.

```{r}
library(infotheo)

mi_mat <- mutinformation(eval_data)
mi_mat

# just the row for the target, sorted so the strongest predictors are
# obvious at a glance
sort(mi_mat[-7, 7], decreasing = TRUE)
```

I used `corrplot` here rather than a plain heatmap - it's normally built for 
correlation matrices but works fine on any square numeric matrix as long as 
`is.corr = FALSE` is set, and I think it reads a bit more clearly than a 
ggplot tile plot for this many variables.

```{r}
library(corrplot)

corrplot(mi_mat, method = "color", is.corr = FALSE,
         col = colorRampPalette(c("#FFFFFF", "#F4A261", "#8B0000"))(100),
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Mutual Information Between Variables", mar = c(0, 0, 2, 0))
```

## 1.5 Chi-squared tests

```{r}
chisq.test(eval_data$buying, eval_data$acceptability)
chisq.test(eval_data$buying, eval_data$maint)

# checking persons vs acceptability too, since the plots above suggested a
# fairly strong relationship there
chisq.test(eval_data$persons, eval_data$acceptability)
```

## 1.6 Train/test split

This bit is given in the assignment brief and needs to stay exactly as-is 
(same seed, same split ratio) so everyone's results are comparable.

```{r}
library(caTools)
set.seed(123)
split <- sample.split(eval_data$acceptability, SplitRatio = 0.8)
part_train <- subset(eval_data, split == TRUE)
part_test  <- subset(eval_data, split == FALSE)

summary(part_train$acceptability)
summary(part_test$acceptability)
prop.table(table(part_train$acceptability))
prop.table(table(part_test$acceptability))
```

---

# Part 2a: Building the Models

### Decision Tree (C5.0)

```{r}
library(caret)
library(C50)

cart_fit  <- C5.0(part_train[-7], part_train$acceptability)
cart_pred <- predict(cart_fit, part_test)
confusionMatrix(cart_pred, part_test$acceptability)
```

### Naive Bayes

```{r}
library(e1071)
set.seed(123)

bayes_fit  <- naiveBayes(part_train[, -7], part_train$acceptability)
bayes_pred <- predict(bayes_fit, part_test)
confusionMatrix(bayes_pred, part_test$acceptability)
```

### Multinomial Logistic Regression

```{r}
library(nnet)

logit_fit  <- multinom(acceptability ~ ., data = part_train)
logit_pred <- predict(logit_fit, part_test)
confusionMatrix(logit_pred, part_test$acceptability)
```

### SVM

```{r}
library(kernlab)

# linear kernel first
svmLin_fit  <- ksvm(acceptability ~ ., data = part_train, kernel = "vanilladot")
svmLin_pred <- predict(svmLin_fit, part_test)
confusionMatrix(svmLin_pred, part_test$acceptability)
```

```{r}
# and the e1071 default (radial kernel) for comparison - curious if the
# non-linear boundary does any better here
svmRad_fit  <- svm(acceptability ~ ., data = part_train)
svmRad_pred <- predict(svmRad_fit, part_test)
confusionMatrix(svmRad_pred, part_test$acceptability)
```

### Neural Network

```{r}
# following the nnet approach from GeeksforGeeks:
# https://www.geeksforgeeks.org/r-language/neural-networks-using-the-r-nnet-package/
mlp_fit  <- nnet(acceptability ~ ., data = part_train, size = 5)
mlp_pred <- predict(mlp_fit, part_test, type = "class")
confusionMatrix(factor(mlp_pred), part_test$acceptability)
```

## Cross-validated versions

Same five algorithms, but this time letting `caret::train()` handle 10-fold 
CV so the accuracy estimate isn't just dependent on one lucky/unlucky split.

### Decision Tree - CV

```{r}
cv_setup <- trainControl(method = "cv", number = 10)
set.seed(123)

cart_cv_fit  <- train(part_train[, -7], part_train[, 7], method = "C5.0", trControl = cv_setup)
cart_cv_fit
cart_cv_pred <- predict(cart_cv_fit, part_test)
confusionMatrix(cart_cv_pred, part_test$acceptability)
```

### Naive Bayes - CV

```{r}
cv_setup <- trainControl(method = "cv", number = 10)
set.seed(123)

bayes_cv_fit  <- train(part_train[, -7], part_train[, 7], method = "nb", trControl = cv_setup)
bayes_cv_fit
bayes_cv_pred <- predict(bayes_cv_fit, part_test)
confusionMatrix(bayes_cv_pred, part_test$acceptability)
```

### Logistic Regression - CV

```{r}
cv_setup <- trainControl(method = "cv", number = 10)
set.seed(123)

logit_cv_fit  <- train(part_train[, -7], part_train[, 7], method = "multinom",
                        trControl = cv_setup, trace = FALSE)
logit_cv_fit
logit_cv_pred <- predict(logit_cv_fit, part_test)
confusionMatrix(logit_cv_pred, part_test$acceptability)
```

### Neural Network - CV

```{r}
cv_setup <- trainControl(method = "cv", number = 10)
set.seed(123)

mlp_cv_fit  <- train(part_train[, -7], part_train[, 7], method = "nnet",
                      trControl = cv_setup, trace = FALSE)
mlp_cv_fit
mlp_cv_pred <- predict(mlp_cv_fit, part_test)
confusionMatrix(mlp_cv_pred, part_test$acceptability)
```

### SVM - CV

```{r}
set.seed(123)
svm_cv_fit  <- ksvm(acceptability ~ ., data = part_train, kernel = "vanilladot", cross = 10)
svm_cv_fit
svm_cv_pred <- predict(svm_cv_fit, part_test)
confusionMatrix(svm_cv_pred, part_test$acceptability)
```

## Feature selection - trying a few different values of x

Rather than just picking the top 3 features and stopping there, I ranked 
all six by their mutual information with the target and then looped 
through the top 2, top 3, top 4 and top 5 to see where accuracy actually 
levels off. Decision tree is used as the test algorithm here since it 
trains quickly.

```{r}
mi_train <- mutinformation(part_train)
ranked_features <- names(sort(mi_train[7, -7], decreasing = TRUE))
ranked_features  # feature names, best predictor first
```

```{r}
# loop over a few candidate values of x and record accuracy for each
feature_counts <- c(2, 3, 4, 5)
fs_results <- data.frame(top_x = integer(), accuracy = numeric())

for (x in feature_counts) {
  chosen_cols <- ranked_features[1:x]
  
  fs_fit  <- C5.0(part_train[, chosen_cols], part_train$acceptability)
  fs_pred <- predict(fs_fit, part_test)
  acc <- confusionMatrix(fs_pred, part_test$acceptability)$overall["Accuracy"]
  
  fs_results <- rbind(fs_results, data.frame(top_x = x, accuracy = acc))
}

fs_results
```

```{r}
# quick plot so it's obvious where extra features stop helping
ggplot(fs_results, aes(x = top_x, y = accuracy)) +
  geom_line(color = "#457B9D", linewidth = 1) +
  geom_point(color = "#1D3557", size = 3) +
  labs(title = "Decision Tree Accuracy vs Number of Selected Features",
       x = "Number of top features used (x)", y = "Accuracy on test set") +
  theme_minimal()
```

Repeating the same idea with Naive Bayes and multinomial logistic 
regression, just using top 3 as a representative case (matching what the 
brief's example used) rather than re-running the full loop for every 
algorithm:

```{r}
top3_cols <- ranked_features[1:3]

bayes_fs_fit  <- naiveBayes(part_train[, top3_cols], part_train$acceptability)
bayes_fs_pred <- predict(bayes_fs_fit, part_test)
confusionMatrix(bayes_fs_pred, part_test$acceptability)
```

```{r}
# multinom needs a formula, so I build one dynamically from the chosen columns
fs_formula <- as.formula(paste("acceptability ~", paste(top3_cols, collapse = " + ")))
logit_fs_fit  <- multinom(fs_formula, data = part_train)
logit_fs_pred <- predict(logit_fs_fit, part_test)
confusionMatrix(logit_fs_pred, part_test$acceptability)
```

---

# Part 2b: Hyperparameter Tuning

## Neural network - decay

I used a wider spread of decay values than a simple 0-to-0.1 sweep, 
including a couple of larger values, to check whether accuracy keeps 
improving or starts dropping again once regularisation gets too strong.

```{r}
tune_setup <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
mlp_grid <- expand.grid(size = c(3, 5, 7),
                         decay = c(0, 0.0005, 0.001, 0.01, 0.05, 0.1, 0.3))
set.seed(123)

mlp_tuned <- train(part_train[, -7], part_train[, 7], method = "nnet",
                    trControl = tune_setup, tuneGrid = mlp_grid, trace = FALSE)
plot(mlp_tuned)
```

## Decision tree - trials

Also widened this grid a bit compared to the lab example, going up to 150 
boosting trials to see where the accuracy curve actually flattens out.

```{r}
tune_setup <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
cart_grid <- expand.grid(model = "tree",
                          trials = c(1, 5, 15, 25, 40, 60, 80, 100),
                          winnow = FALSE)
set.seed(123)

cart_tuned <- train(part_train[, -7], part_train[, 7], method = "C5.0",
                     trControl = tune_setup, tuneGrid = cart_grid)
plot(cart_tuned)
```

---

# Part 3. Presentation of Specified Results

*(To be finalised in the lab session once the exact figures required are confirmed - the feature-selection accuracy-vs-x plot above and the two tuning plots are strong candidates to include here.)*

## References

- Kuhn, M. (2008) 'Building predictive models in R using the caret package', 
  *Journal of Statistical Software*, 28(5), pp. 1-26.
- Wei, T. and Simko, V. (2021) *corrplot: Visualization of a Correlation 
  Matrix*. R package.
- UCI Machine Learning Repository (1997) *Car Evaluation Data Set*. Available 
  at: https://archive.ics.uci.edu/dataset/19/car+evaluation (Accessed: [add date]).
- Wickham, H. (2016) *ggplot2: Elegant Graphics for Data Analysis*. New York: Springer-Verlag.
