---
title: "CMP330 Data Analytics - Assessment 2 (Pre-Assignment)"
author: "MD Ali Hasan Jidan (B01011849)"
output:
  html_notebook: default
  pdf_document: default
---

# What this notebook does

I'm building a model that looks at a car's price, maintenance cost, doors, 
seats, boot size and safety, and predicts whether it counts as unacceptable, 
acceptable, good, or very good. I'll compare five different algorithms to 
see which works best.

---

# Getting the data ready

## Reading and cleaning

```{r}
vehicle.df <- read.csv("car.csv")
vehicle.df <- data.frame(lapply(vehicle.df, factor))

head(vehicle.df)
str(vehicle.df)
```

## Looking at the raw numbers first

```{r}
# How many cars per acceptability class
table(vehicle.df$acceptability)

# Buying price against acceptability
table(vehicle.df$buying, vehicle.df$acceptability)

# Extra cross-tab: maintenance cost against safety - checking if 
# expensive-to-maintain cars tend to also be less safe
table(vehicle.df$maint, vehicle.df$safety)

# Extra cross-tab: doors against acceptability
table(vehicle.df$doors, vehicle.df$acceptability)
```

## Charts

```{r}
library(ggplot2)

vehicle.df$acceptability <- factor(vehicle.df$acceptability,
                                    levels = c("unacc", "acc", "good", "vgood"))

# Overall distribution
ggplot(vehicle.df, aes(x = acceptability)) +
  geom_bar(fill = "#556B2F") +
  labs(title = "Overall spread of car acceptability", x = "Acceptability", y = "Count") +
  theme_bw()

# Split by buying price
ggplot(vehicle.df, aes(x = acceptability, fill = buying)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("#E76F51", "#F4A261", "#2A9D8F", "#264653")) +
  labs(title = "Acceptability split by buying price", x = "Acceptability", y = "Count") +
  theme_bw()

# Extra chart: split by lug_boot size
ggplot(vehicle.df, aes(x = acceptability, fill = lug_boot)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Accent") +
  labs(title = "Acceptability split by boot size", x = "Acceptability", y = "Count") +
  theme_bw()
```

## Mutual information instead of correlation

Correlation only makes sense for numeric data, and everything here is 
categorical, so mutual information is the right tool instead.

```{r}
library(infotheo)

mi.table <- mutinformation(vehicle.df)
mi.table

# How informative is each predictor about the target column?
mi.table[-7, 7]
```

```{r}
library(reshape2)

mi.long <- melt(mi.table)
ggplot(mi.long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#FDE0DD", high = "#980043") +
  labs(title = "Mutual information heatmap across all variables", x = "", y = "", fill = "MI") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Chi-squared tests

```{r}
chisq.test(vehicle.df$buying, vehicle.df$acceptability)
chisq.test(vehicle.df$buying, vehicle.df$maint)

# Extra test: persons against acceptability
chisq.test(vehicle.df$persons, vehicle.df$acceptability)
```

## Splitting the data

```{r}
library(caTools)
set.seed(123)
split <- sample.split(vehicle.df$acceptability, SplitRatio = 0.8)
vehicle.train <- subset(vehicle.df, split == TRUE)
vehicle.test  <- subset(vehicle.df, split == FALSE)

summary(vehicle.train$acceptability)
summary(vehicle.test$acceptability)
prop.table(table(vehicle.train$acceptability))
prop.table(table(vehicle.test$acceptability))
```

---

# Building and testing the models

### Decision tree (C5.0)

```{r}
library(caret)
library(C50)

c50.fit  <- C5.0(vehicle.train[-7], vehicle.train$acceptability)
c50.pred <- predict(c50.fit, vehicle.test)
confusionMatrix(c50.pred, vehicle.test$acceptability)
```

### Naive Bayes

```{r}
library(e1071)
set.seed(123)

nb.fit  <- naiveBayes(vehicle.train[, -7], vehicle.train$acceptability)
nb.pred <- predict(nb.fit, vehicle.test)
confusionMatrix(nb.pred, vehicle.test$acceptability)
```

### Logistic regression (multinomial)

```{r}
library(nnet)

mlr.fit  <- multinom(acceptability ~ ., data = vehicle.train)
mlr.pred <- predict(mlr.fit, vehicle.test)
confusionMatrix(mlr.pred, vehicle.test$acceptability)
```

### SVM

```{r}
library(kernlab)

svm.fit.linear  <- ksvm(acceptability ~ ., data = vehicle.train, kernel = "vanilladot")
svm.pred.linear <- predict(svm.fit.linear, vehicle.test)
confusionMatrix(svm.pred.linear, vehicle.test$acceptability)
```

```{r}
svm.fit.default  <- svm(acceptability ~ ., data = vehicle.train)
svm.pred.default <- predict(svm.fit.default, vehicle.test)
confusionMatrix(svm.pred.default, vehicle.test$acceptability)
```

### Neural network

```{r}
# Reference: https://www.geeksforgeeks.org/r-language/neural-networks-using-the-r-nnet-package/
ann.fit  <- nnet(acceptability ~ ., data = vehicle.train, size = 5)
ann.pred <- predict(ann.fit, vehicle.test, type = "class")
confusionMatrix(factor(ann.pred), vehicle.test$acceptability)
```

## Same models, but with 10-fold cross-validation

### Decision tree, cross-validated

```{r}
ctrl.cv <- trainControl(method = "cv", number = 10)
set.seed(123)

c50.cv.fit  <- train(vehicle.train[, -7], vehicle.train[, 7], method = "C5.0", trControl = ctrl.cv)
c50.cv.fit
c50.cv.pred <- predict(c50.cv.fit, vehicle.test)
confusionMatrix(c50.cv.pred, vehicle.test$acceptability)
```

### Naive Bayes, cross-validated

```{r}
ctrl.cv <- trainControl(method = "cv", number = 10)
set.seed(123)

nb.cv.fit  <- train(vehicle.train[, -7], vehicle.train[, 7], method = "nb", trControl = ctrl.cv)
nb.cv.fit
nb.cv.pred <- predict(nb.cv.fit, vehicle.test)
confusionMatrix(nb.cv.pred, vehicle.test$acceptability)
```

### Logistic regression, cross-validated

```{r}
ctrl.cv <- trainControl(method = "cv", number = 10)
set.seed(123)

mlr.cv.fit  <- train(vehicle.train[, -7], vehicle.train[, 7], method = "multinom",
                      trControl = ctrl.cv, trace = FALSE)
mlr.cv.fit
mlr.cv.pred <- predict(mlr.cv.fit, vehicle.test)
confusionMatrix(mlr.cv.pred, vehicle.test$acceptability)
```

### Neural network, cross-validated

```{r}
ctrl.cv <- trainControl(method = "cv", number = 10)
set.seed(123)

ann.cv.fit  <- train(vehicle.train[, -7], vehicle.train[, 7], method = "nnet",
                      trControl = ctrl.cv, trace = FALSE)
ann.cv.fit
ann.cv.pred <- predict(ann.cv.fit, vehicle.test)
confusionMatrix(ann.cv.pred, vehicle.test$acceptability)
```

### SVM, cross-validated

```{r}
set.seed(123)
svm.cv.fit  <- ksvm(acceptability ~ ., data = vehicle.train, kernel = "vanilladot", cross = 10)
svm.cv.fit
svm.cv.pred <- predict(svm.cv.fit, vehicle.test)
confusionMatrix(svm.cv.pred, vehicle.test$acceptability)
```

## Trying fewer, more informative features

```{r}
mi.train <- mutinformation(vehicle.train)
mi.train
mi.train[7, -7]
```

`persons`, `lug_boot`, and `safety` come out on top, so those three are used 
below instead of all six features.

### Decision tree with top features only

```{r}
c50.fs.fit  <- C5.0(vehicle.train[, 4:6], vehicle.train$acceptability)
c50.fs.pred <- predict(c50.fs.fit, vehicle.test)
confusionMatrix(c50.fs.pred, vehicle.test$acceptability)
```

### Naive Bayes with top features only

```{r}
nb.fs.fit  <- naiveBayes(vehicle.train[, 4:6], vehicle.train$acceptability)
nb.fs.pred <- predict(nb.fs.fit, vehicle.test)
confusionMatrix(nb.fs.pred, vehicle.test$acceptability)
```

### Logistic regression with top features only

```{r}
mlr.fs.fit  <- multinom(acceptability ~ persons + lug_boot + safety, data = vehicle.train)
mlr.fs.pred <- predict(mlr.fs.fit, vehicle.test)
confusionMatrix(mlr.fs.pred, vehicle.test$acceptability)
```

---

# Fine-tuning for better accuracy

## Neural network - tuning decay

```{r}
tune.ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
ann.grid  <- expand.grid(size = 5, decay = c(0, 0.001, 0.005, 0.01, 0.02, 0.05, 0.1))
set.seed(123)

ann.tuned <- train(vehicle.train[, -7], vehicle.train[, 7], method = "nnet",
                    trControl = tune.ctrl, tuneGrid = ann.grid, trace = FALSE)
plot(ann.tuned)
```

## Decision tree - tuning number of boosting trials

```{r}
tune.ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
c50.grid  <- expand.grid(model = "tree",
                          trials = c(1, 5, 10, 15, 20, 25, 30, 40, 50, 60, 80, 100),
                          winnow = FALSE)
set.seed(123)

c50.tuned <- train(vehicle.train[, -7], vehicle.train[, 7], method = "C5.0",
                    trControl = tune.ctrl, tuneGrid = c50.grid)
plot(c50.tuned)
```

---

# Part 3. Presentation of Specified Results

*(To be finalised in the lab session once exact required figures are confirmed.)*

## References

- Kuhn, M. (2008) 'Building predictive models in R using the caret package', 
  *Journal of Statistical Software*, 28(5), pp. 1-26.
- UCI Machine Learning Repository (1997) *Car Evaluation Data Set*. Available 
  at: https://archive.ics.uci.edu/dataset/19/car+evaluation (Accessed: [add date]).
- Wickham, H. (2016) *ggplot2: Elegant Graphics for Data Analysis*. New York: Springer-Verlag.
