---
title: "CMP330 Assessment 2"
author: "Topu"
output: html_notebook
---
This is code provides solutions for the pre-assignment and follow-up lab test.  
First, read in the file and convert everything to **factors**. This code will be shared with students 
in the specification for the pre-assignment that is released to students.
```{r} 
df = read.csv("car.csv") 
df = data.frame(lapply(df, factor)) 
head(df) 
str(df) 
``` 
# 1. Exploratory data analysis and data preparation 
General information about target and other variables. 
```{r} 
table(df$acceptability) 
table(df$buying,df$acceptability) 
summary(df) 
``` 
Plot distributions of variables. 
```{r} 
library(ggplot2) 
df$acceptability = factor(df$acceptability, levels = c("unacc","acc","good","vgood")) 
ggplot(df,aes(x=acceptability)) + geom_bar() 
ggplot(df,aes(x=acceptability, fill=buying)) + geom_bar() 
``` 
Mutual information.


```{r} 
#install.packages("infotheo") 
library("infotheo") 
m = mutinformation(df) 
m 
#Could use following code for each row, replacing [-1,1] with [-2,2], etc. 
max(m[-1,1]) 
``` 
```{r} 
chisq.test(df$buying,df$acceptability) 
chisq.test(df$buying,df$maint) 
``` 
Split data into training and test sets. 
```{r} 
library(caTools) 
set.seed(123) 
split = sample.split(df$acceptability, SplitRatio = 0.8) 
df_train = subset(df, split == TRUE) 
df_test = subset(df, split == FALSE) 
summary(df_train$acceptability) 
summary(df_test$acceptability) 
prop.table(table(df_train$acceptability)) 
prop.table(table(df_test$acceptability)) 
``` 
# 2a. Data Analytics solution: Application of machine learning 
### Decision Tree, C5.0 
```{r} 
library(caret) 
library(C50) 
dt_model <- C5.0(df_train[-7], df_train$acceptability) 
dt_pred <- predict(dt_model, df_test) 
confusionMatrix(dt_pred, df_test$acceptability) 
``` 
### Naive Bayes 
```{r} 
library(e1071) 
set.seed(123) 
nb_model = naiveBayes(df_train[,-7],df_train$acceptability)  
nb_pred = predict(nb_model,df_test) 
confusionMatrix(nb_pred, df_test$acceptability) 
``` 
### Logistic Regression 
```{r} 
library(nnet) 
lg_model <- multinom(acceptability ~ ., data = df_train) 
lg_pred <-predict(lg_model, df_test) 
confusionMatrix(lg_pred, df_test$acceptability) 
``` 
### SVM 
```{r} 
library(kernlab) 
svm_model <- ksvm(acceptability ~ ., data = df_train, kernel = "vanilladot") 
svm_pred <- predict(svm_model, df_test) 
confusionMatrix(svm_pred, df_test$acceptability) 
``` 
```{r} 
library(e1071) 
svm_model <- svm(acceptability ~ ., data = df_train) 
svm_pred <- predict(svm_model, df_test) 
confusionMatrix(svm_pred, df_test$acceptability) 
``` 
### Neural Network 
```{r} 
# see https://www.geeksforgeeks.org/r-language/neural-networks-using-the-r-nnet-package/ 
nn_model <- nnet(acceptability ~ ., data = df_train, size = 5) 
nn_pred <-predict(nn_model, df_test, type="class") 
confusionMatrix(factor(nn_pred), df_test$acceptability) 
``` 
### Decision Tree with CV 
```{r} 
ctrl = trainControl(method='cv',number=10) 
set.seed(123) 
dt_cv_model = train(df_train[,-7],df_train[,7],'C5.0',trControl=ctrl) 
dt_cv_model 
dt_cv_pred = predict(dt_cv_model,df_test) 
confusionMatrix(dt_cv_pred, df_test$acceptability) 
``` 
### Naive Bayes with CV 
```{r} 
ctrl = trainControl(method='cv',number=10) 
set.seed(123) 
nb_cv_model = train(df_train[,-7],df_train[,7],'nb',trControl=ctrl) 
nb_cv_model 
nb_cv_pred = predict(nb_cv_model,df_test) 
confusionMatrix(nb_cv_pred, df_test$acceptability) 
``` 
### Logistic Regression with CV 
```{r} 
ctrl = trainControl(method='cv',number=10) 
set.seed(123) 
lg_cv_model = train(df_train[,-7],df_train[,7],'multinom',trControl=ctrl, trace=FALSE) 
lg_cv_model 
lg_cv_pred = predict(lg_cv_model,df_test) 
confusionMatrix(lg_cv_pred, df_test$acceptability) 
``` 
### Neural Network with CV 
```{r} 
ctrl = trainControl(method='cv',number=10) 
set.seed(123) 
nn_cv_model = train(df_train[,-7],df_train[,7],'nnet',trControl=ctrl,trace=FALSE) 
nn_cv_model 
nn_cv_pred = predict(nn_cv_model,df_test) 
confusionMatrix(nn_cv_pred, df_test$acceptability) 
``` 
### SVM with CV 
```{r} 
ctrl = trainControl(method='cv',number=10) 
set.seed(123) 
svm_cv_model <- ksvm(acceptability ~ ., data = df_train, kernel = "vanilladot", cross=10) 
svm_cv_model 
svm_cv_pred <- predict(svm_cv_model, df_test) 
confusionMatrix(svm_cv_pred, df_test$acceptability) 
``` 
### Feature Selection 
```{r} 
m = mutinformation(df_train) 
m 
m[7,-7] 
``` 
### Decision Tree with Feature Selection 
```{r} 
dt_fs_model <- C5.0(df_train[,4:6], df_train$acceptability) 
dt_fs_pred <- predict(dt_fs_model, df_test) 
confusionMatrix(dt_fs_pred, df_test$acceptability) 
``` 
### Naive Bayes with Feature Selection 
```{r} 
nb_fs_model = naiveBayes(df_train[,4:6],df_train$acceptability) 
nb_fs_pred = predict(nb_fs_model,df_test) 
confusionMatrix(nb_fs_pred, df_test$acceptability) 
``` 
### Logistic Regression with Feature Selection 
```{r} 
lg_fs_model <- multinom(acceptability ~ persons + lug_boot + safety, data = df_train) 
lg_fs_pred <-predict(lg_fs_model, df_test) 
confusionMatrix(lg_fs_pred, df_test$acceptability) 
``` 
# 2b. Data Analytics solution: Improving model performance 
### Tuning for Neural Network 
```{r} 
ctrl = trainControl(method="repeatedcv",number=10,repeats=10) 
grid <- expand.grid(size=5,decay=c(0,0.001,0.005,0.01,0.02,0.05,0.1)) 
set.seed(123) 
m <- train(df_train[,-7],df_train[,7], method = "nnet", trControl = ctrl,tuneGrid = 
grid,trace=FALSE) 
plot(m) 
``` 
### Tuning for DT 
```{r} 
ctrl = trainControl(method="repeatedcv",number=10,repeats=10) 
grid <- expand.grid(model="tree",trials=c(1,5,10,15,20,25,30,40,50,60,80,100),winnow=FALSE) 
set.seed(123) 
m <- train(df_train[,-7],df_train[,7], method = "C5.0", trControl = ctrl,tuneGrid = grid) 
plot(m) 
```