---
title: "CMP330 Data Analytics - Assessment 2 (Pre-Assignment)"
author: 
output:
  html_notebook: default
  pdf_document: default
---

# What am I trying to do here?

In this notebook I'm going to predict how "acceptable" a car is 
(`unacc`, `acc`, `good`, `vgood`) using six features about the car - things 
like buying price, maintenance cost, number of doors, seats, boot size, and 
safety rating. I'll try five different ML algorithms and compare them.

---

# Step 1: Getting to know the data

## 1.1 Reading in the dataset

```{r}
car_records <- read.csv("car.csv")
car_records <- data.frame(lapply(car_records, factor))

head(car_records)
str(car_records)
```

## 1.2 A first look at the numbers

Before I build any models I want to know roughly how the classes are spread 
out, and whether some obvious features look related to acceptability.

```{r}
# How many cars in each acceptability bucket?
table(car_records$acceptability)

# Buying price vs acceptability
table(car_records$buying, car_records$acceptability)

# I also checked maintenance cost vs acceptability, out of curiosity
table(car_records$maint, car_records$acceptability)

# And number of doors vs persons - just to see if there's an obvious link
table(car_records$doors, car_records$persons)
```

## 1.3 Putting it into pictures

```{r}
library(ggplot2)

car_records$acceptability <- factor(car_records$acceptability,
                                     levels = c("unacc", "acc", "good", "vgood"))

# Overall spread of acceptability
ggplot(car_records, aes(x = acceptability)) +
  geom_bar(fill = "darkorange") +
  labs(title = "How acceptable are the cars overall?",
       x = "Acceptability", y = "Number of cars") +
  theme_minimal()

# Acceptability split by buying price
ggplot(car_records, aes(x = acceptability, fill = buying)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Does buying price affect acceptability?",
       x = "Acceptability", y = "Number of cars") +
  theme_minimal()

# Extra: acceptability split by number of persons the car seats
ggplot(car_records, aes(x = acceptability, fill = persons)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("#FF6B6B", "#4ECDC4", "#1A535C")) +
  labs(title = "Does seating capacity affect acceptability?",
       x = "Acceptability", y = "Number of cars") +
  theme_minimal()
```

## 1.4 Checking how much each feature "tells us" about the others

Regular correlation doesn't work here because everything is categorical, so 
I used mutual information instead - it tells us how much knowing one 
variable reduces our uncertainty about another.

```{r}
library(infotheo)

info_scores <- mutinformation(car_records)
info_scores

# How informative is each feature about acceptability specifically?
info_scores[-7, 7]
```

```{r}
library(reshape2)

info_long <- melt(info_scores)
ggplot(info_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#FFF7EC", high = "#7A0177") +
  labs(title = "Mutual information between all variables", x = "", y = "", fill = "MI score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 1.5 Testing independence with chi-squared

```{r}
chisq.test(car_records$buying, car_records$acceptability)
chisq.test(car_records$buying, car_records$maint)

# Extra check - is boot size independent of acceptability?
chisq.test(car_records$lug_boot, car_records$acceptability)
```

## 1.6 Splitting into training and testing data

This part has to stay exactly as given in the brief so results line up 
across the class.

```{r}
library(caTools)
set.seed(123)
split <- sample.split(car_records$acceptability, SplitRatio = 0.8)
train_set <- subset(car_records, split == TRUE)
test_set  <- subset(car_records, split == FALSE)

summary(train_set$acceptability)
summary(test_set$acceptability)
prop.table(table(train_set$acceptability))
prop.table(table(test_set$acceptability))
```

---

# Step 2a: Trying out different ML algorithms

### Decision tree (C5.0)

```{r}
library(caret)
library(C50)

tree_fit  <- C5.0(train_set[-7], train_set$acceptability)
tree_pred <- predict(tree_fit, test_set)
confusionMatrix(tree_pred, test_set$acceptability)
```

### Naive Bayes

```{r}
library(e1071)
set.seed(123)

bayes_fit  <- naiveBayes(train_set[, -7], train_set$acceptability)
bayes_pred <- predict(bayes_fit, test_set)
confusionMatrix(bayes_pred, test_set$acceptability)
```

### Multinomial logistic regression

```{r}
library(nnet)

logit_fit  <- multinom(acceptability ~ ., data = train_set)
logit_pred <- predict(logit_fit, test_set)
confusionMatrix(logit_pred, test_set$acceptability)
```

### Support Vector Machine

```{r}
library(kernlab)

svm_fit_a  <- ksvm(acceptability ~ ., data = train_set, kernel = "vanilladot")
svm_pred_a <- predict(svm_fit_a, test_set)
confusionMatrix(svm_pred_a, test_set$acceptability)
```

```{r}
# Trying the e1071 SVM implementation as well, just to compare
svm_fit_b  <- svm(acceptability ~ ., data = train_set)
svm_pred_b <- predict(svm_fit_b, test_set)
confusionMatrix(svm_pred_b, test_set$acceptability)
```

### Neural network

```{r}
# Reference: https://www.geeksforgeeks.org/r-language/neural-networks-using-the-r-nnet-package/
net_fit  <- nnet(acceptability ~ ., data = train_set, size = 5)
net_pred <- predict(net_fit, test_set, type = "class")
confusionMatrix(factor(net_pred), test_set$acceptability)
```

## Now with cross-validation via caret

### Decision tree with CV

```{r}
cv_ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

tree_fit_cv  <- train(train_set[, -7], train_set[, 7], method = "C5.0", trControl = cv_ctrl)
tree_fit_cv
tree_pred_cv <- predict(tree_fit_cv, test_set)
confusionMatrix(tree_pred_cv, test_set$acceptability)
```

### Naive Bayes with CV

```{r}
cv_ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

bayes_fit_cv  <- train(train_set[, -7], train_set[, 7], method = "nb", trControl = cv_ctrl)
bayes_fit_cv
bayes_pred_cv <- predict(bayes_fit_cv, test_set)
confusionMatrix(bayes_pred_cv, test_set$acceptability)
```

### Logistic regression with CV

```{r}
cv_ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

logit_fit_cv  <- train(train_set[, -7], train_set[, 7], method = "multinom",
                        trControl = cv_ctrl, trace = FALSE)
logit_fit_cv
logit_pred_cv <- predict(logit_fit_cv, test_set)
confusionMatrix(logit_pred_cv, test_set$acceptability)
```

### Neural network with CV

```{r}
cv_ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

net_fit_cv  <- train(train_set[, -7], train_set[, 7], method = "nnet",
                      trControl = cv_ctrl, trace = FALSE)
net_fit_cv
net_pred_cv <- predict(net_fit_cv, test_set)
confusionMatrix(net_pred_cv, test_set$acceptability)
```

### SVM with CV

```{r}
set.seed(123)
svm_fit_cv  <- ksvm(acceptability ~ ., data = train_set, kernel = "vanilladot", cross = 10)
svm_fit_cv
svm_pred_cv <- predict(svm_fit_cv, test_set)
confusionMatrix(svm_pred_cv, test_set$acceptability)
```

## Picking the best features using mutual information

```{r}
info_train <- mutinformation(train_set)
info_train
info_train[7, -7]
```

I'm going with the top three features - `persons`, `lug_boot`, `safety` - 
based on their mutual information scores with the target.

### Decision tree with fewer features

```{r}
tree_fit_top3  <- C5.0(train_set[, 4:6], train_set$acceptability)
tree_pred_top3 <- predict(tree_fit_top3, test_set)
confusionMatrix(tree_pred_top3, test_set$acceptability)
```

### Naive Bayes with fewer features

```{r}
bayes_fit_top3  <- naiveBayes(train_set[, 4:6], train_set$acceptability)
bayes_pred_top3 <- predict(bayes_fit_top3, test_set)
confusionMatrix(bayes_pred_top3, test_set$acceptability)
```

### Logistic regression with fewer features

```{r}
logit_fit_top3  <- multinom(acceptability ~ persons + lug_boot + safety, data = train_set)
logit_pred_top3 <- predict(logit_fit_top3, test_set)
confusionMatrix(logit_pred_top3, test_set$acceptability)
```

---

# Step 2b: Squeezing out more performance

## Tuning the neural network's decay rate

```{r}
tune_ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
net_grid  <- expand.grid(size = 5, decay = c(0, 0.001, 0.005, 0.01, 0.02, 0.05, 0.1))
set.seed(123)

net_tuned <- train(train_set[, -7], train_set[, 7], method = "nnet",
                    trControl = tune_ctrl, tuneGrid = net_grid, trace = FALSE)
plot(net_tuned)
```

## Tuning the decision tree's number of trials

```{r}
tune_ctrl  <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
tree_grid  <- expand.grid(model = "tree",
                           trials = c(1, 5, 10, 15, 20, 25, 30, 40, 50, 60, 80, 100),
                           winnow = FALSE)
set.seed(123)

tree_tuned <- train(train_set[, -7], train_set[, 7], method = "C5.0",
                     trControl = tune_ctrl, tuneGrid = tree_grid)
plot(tree_tuned)
```

---

# Part 3. Presentation of Specified Results

*(To be finalised in the lab once the exact figures required are confirmed.)*

## References

- Kuhn, M. (2008) 'Building predictive models in R using the caret package', 
  *Journal of Statistical Software*, 28(5), pp. 1-26.
- UCI Machine Learning Repository (1997) *Car Evaluation Data Set*. Available 
  at: https://archive.ics.uci.edu/dataset/19/car+evaluation (Accessed: [add date]).
- Wickham, H. (2016) *ggplot2: Elegant Graphics for Data Analysis*. New York: Springer-Verlag.
