---
title: "CMP330 Data Analytics - Assessment 2 (Pre-Assignment)"
author: "MD Ali Hasan Jidan (B01011849)"
output:
  html_notebook: default
  pdf_document: default
---

# Introduction

This notebook contains my solution to the CMP330 Assessment 2 pre-assignment. 
The goal is to classify cars according to their **acceptability** 
(`unacc`, `acc`, `good`, `vgood`) based on six categorical features 
(`buying`, `maint`, `doors`, `persons`, `lug_boot`, `safety`) using several 
machine learning algorithms.

---

# Part 1: Exploratory Data Analysis and Data Preparation

## 1.1 Loading the data

The dataset is read in and every column is converted to a factor, since all 
variables (including the target) are categorical rather than numeric.

```{r}
car_data <- read.csv("car.csv")
car_data <- data.frame(lapply(car_data, factor))

head(car_data)
str(car_data)
```

## 1.2 Basic frequency counts

Before modelling, it's useful to see how the classes are distributed and how 
individual features relate to the target label.

```{r}
# How many cars fall into each acceptability class?
table(car_data$acceptability)

# Cross-tabulation: buying price vs acceptability
table(car_data$buying, car_data$acceptability)

# Cross-tabulation: safety vs acceptability (an extra one I added, since 
# safety is intuitively a strong predictor of acceptability)
table(car_data$safety, car_data$acceptability)

summary(car_data)
```

## 1.3 Visualising the distributions

I used `ggplot2` to visualise the target distribution on its own, and then 
broken down by other features to get a first impression of which variables 
might be useful predictors.

```{r}
library(ggplot2)

# Fix the ordering of the acceptability levels so charts read logically
car_data$acceptability <- factor(car_data$acceptability,
                                  levels = c("unacc", "acc", "good", "vgood"))

# Overall distribution of the target class
ggplot(car_data, aes(x = acceptability)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Car Acceptability",
       x = "Acceptability", y = "Count")
```
*Figure 1: Most cars in the dataset are rated unacc (693 out of 1000), 
while only a small number are good or vgood. This shows the classes are 
quite imbalanced, which is worth keeping in mind later when looking at 
model accuracy.*


```{r}
# Acceptability broken down by buying price
ggplot(car_data, aes(x = acceptability, fill = buying)) +
  geom_bar(position = "dodge") +
  labs(title = "Acceptability by Buying Price",
       x = "Acceptability", y = "Count")
```
*Figure 2: Cars with a high or vhigh buying price are almost never rated 
good or vgood - all of those ratings come from cars with low or med buying 
price. This suggests buying price on its own is a decent predictor of 
acceptability, though clearly not the only factor.*

```{r}
# Extra plot: acceptability broken down by safety level
ggplot(car_data, aes(x = acceptability, fill = safety)) +
  geom_bar(position = "dodge") +
  labs(title = "Acceptability by Safety Level",
       x = "Acceptability", y = "Count")
```
*Figure 3: Cars with low safety are never rated above unacc, and vgood 
only shows up for cars with high safety. This makes safety look like one 
of the strongest predictors of acceptability out of all the features, 
which matches what the mutual information scores show later on.*


## 1.4 Mutual information between variables

Since none of the variables are numeric, Pearson correlation isn't 
appropriate. Instead, I used **mutual information** from the `infotheo` 
package, which measures how much knowing one categorical variable reduces 
uncertainty about another.

```{r}
# install.packages("infotheo")
library(infotheo)

mi_matrix <- mutinformation(car_data)
mi_matrix

# Mutual information of every feature with the target (acceptability is 
# column 7)
mi_matrix[-7, 7]
```

I also visualised this as a simple heatmap so the strongest relationships 
are easier to spot at a glance:

```{r}
library(reshape2)

mi_melted <- melt(mi_matrix)
ggplot(mi_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Mutual Information Heatmap", x = "", y = "", fill = "MI") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 1.5 Chi-squared tests of independence

A chi-squared test checks whether two categorical variables are statistically 
independent.

```{r}
chisq.test(car_data$buying, car_data$acceptability)
chisq.test(car_data$buying, car_data$maint)

# Extra test: is safety independent of acceptability?
chisq.test(car_data$safety, car_data$acceptability)
```

## 1.6 Train/test split

As specified in the assignment brief, the data is split 80/20 using a fixed 
seed so that results are reproducible and comparable across students. 
**This exact code must be used as given.**

```{r}
library(caTools)
set.seed(123)
split <- sample.split(car_data$acceptability, SplitRatio = 0.8)
car_train <- subset(car_data, split == TRUE)
car_test  <- subset(car_data, split == FALSE)

summary(car_train$acceptability)
summary(car_test$acceptability)
prop.table(table(car_train$acceptability))
prop.table(table(car_test$acceptability))
```

---

# Part 2a: Application of Machine Learning Algorithms

Each of the five required algorithms is trained on `car_train` and evaluated 
on `car_test` using a confusion matrix (accuracy, kappa, per-class 
sensitivity/specificity, etc.).

### Decision Tree (C5.0)

```{r}
library(caret)
library(C50)

tree_model <- C5.0(car_train[-7], car_train$acceptability)
tree_pred  <- predict(tree_model, car_test)
confusionMatrix(tree_pred, car_test$acceptability)
```

### Naive Bayes

```{r}
library(e1071)
set.seed(123)

nb_model <- naiveBayes(car_train[, -7], car_train$acceptability)
nb_pred  <- predict(nb_model, car_test)
confusionMatrix(nb_pred, car_test$acceptability)
```

### Multinomial Logistic Regression

```{r}
library(nnet)

logreg_model <- multinom(acceptability ~ ., data = car_train)
logreg_pred  <- predict(logreg_model, car_test)
confusionMatrix(logreg_pred, car_test$acceptability)
```

### Support Vector Machine

```{r}
library(kernlab)

svm_model_ksvm <- ksvm(acceptability ~ ., data = car_train, kernel = "vanilladot")
svm_pred_ksvm  <- predict(svm_model_ksvm, car_test)
confusionMatrix(svm_pred_ksvm, car_test$acceptability)
```

```{r}
# Alternative SVM implementation using e1071, for comparison
svm_model_e1071 <- svm(acceptability ~ ., data = car_train)
svm_pred_e1071  <- predict(svm_model_e1071, car_test)
confusionMatrix(svm_pred_e1071, car_test$acceptability)
```

### Neural Network

```{r}
# Reference: https://www.geeksforgeeks.org/r-language/neural-networks-using-the-r-nnet-package/
nn_model <- nnet(acceptability ~ ., data = car_train, size = 5)
nn_pred  <- predict(nn_model, car_test, type = "class")
confusionMatrix(factor(nn_pred), car_test$acceptability)
```

## Using caret::train() with cross-validation

For each algorithm, 10-fold cross-validation is used to tune hyperparameters 
automatically via the `train()` function from `caret`.

### Decision Tree with CV

```{r}
ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

tree_cv_model <- train(car_train[, -7], car_train[, 7], method = "C5.0", trControl = ctrl)
tree_cv_model
tree_cv_pred <- predict(tree_cv_model, car_test)
confusionMatrix(tree_cv_pred, car_test$acceptability)
```

### Naive Bayes with CV

```{r}
ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

nb_cv_model <- train(car_train[, -7], car_train[, 7], method = "nb", trControl = ctrl)
nb_cv_model
nb_cv_pred <- predict(nb_cv_model, car_test)
confusionMatrix(nb_cv_pred, car_test$acceptability)
```

### Logistic Regression with CV

```{r}
ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

logreg_cv_model <- train(car_train[, -7], car_train[, 7], method = "multinom",
                          trControl = ctrl, trace = FALSE)
logreg_cv_model
logreg_cv_pred <- predict(logreg_cv_model, car_test)
confusionMatrix(logreg_cv_pred, car_test$acceptability)
```

### Neural Network with CV

```{r}
ctrl <- trainControl(method = "cv", number = 10)
set.seed(123)

nn_cv_model <- train(car_train[, -7], car_train[, 7], method = "nnet",
                      trControl = ctrl, trace = FALSE)
nn_cv_model
nn_cv_pred <- predict(nn_cv_model, car_test)
confusionMatrix(nn_cv_pred, car_test$acceptability)
```

### SVM with CV

```{r}
set.seed(123)
svm_cv_model <- ksvm(acceptability ~ ., data = car_train, kernel = "vanilladot", cross = 10)
svm_cv_model
svm_cv_pred <- predict(svm_cv_model, car_test)
confusionMatrix(svm_cv_pred, car_test$acceptability)
```

## Feature selection using mutual information

The mutual information between each feature and the target (computed on the 
**training set only**, to avoid data leakage) is used to rank features. The 
top three — `persons`, `lug_boot`, `safety` — are then used to retrain 
simplified models.

```{r}
mi_train <- mutinformation(car_train)
mi_train
mi_train[7, -7]
```
### Trying a range of feature-set sizes (x)

Rather than fixing x = 3, the number of top features is varied across 
x = 2, 3, 4, 5 (ranked by mutual information with the target) and the 
Decision Tree (C5.0) is retrained each time, so the effect of feature-set 
size on accuracy can be compared directly.

```{r}
mi_ranked <- sort(mi_train[7, -7], decreasing = TRUE)
mi_ranked

x_values <- 2:5
fs_accuracy <- data.frame(x = x_values, Accuracy = NA)

for (i in seq_along(x_values)) {
  x <- x_values[i]
  top_features <- names(mi_ranked)[1:x]

  tree_x_model <- C5.0(car_train[, top_features], car_train$acceptability)
  tree_x_pred  <- predict(tree_x_model, car_test)
  acc <- confusionMatrix(tree_x_pred, car_test$acceptability)$overall["Accuracy"]

  fs_accuracy$Accuracy[i] <- acc
  cat("x =", x, "- features:", paste(top_features, collapse = ", "),
      "- Accuracy:", round(acc, 4), "\n")
}

fs_accuracy

ggplot(fs_accuracy, aes(x = x, y = Accuracy)) +
  geom_line(color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  scale_x_continuous(breaks = x_values) +
  labs(title = "Decision Tree accuracy vs. number of top features (x)",
       x = "Number of top features (x)", y = "Test set accuracy")
```

### Decision Tree with selected features

```{r}
tree_fs_model <- C5.0(car_train[, 4:6], car_train$acceptability)
tree_fs_pred  <- predict(tree_fs_model, car_test)
confusionMatrix(tree_fs_pred, car_test$acceptability)
```

### Naive Bayes with selected features

```{r}
nb_fs_model <- naiveBayes(car_train[, 4:6], car_train$acceptability)
nb_fs_pred  <- predict(nb_fs_model, car_test)
confusionMatrix(nb_fs_pred, car_test$acceptability)
```

### Logistic Regression with selected features

```{r}
logreg_fs_model <- multinom(acceptability ~ persons + lug_boot + safety, data = car_train)
logreg_fs_pred  <- predict(logreg_fs_model, car_test)
confusionMatrix(logreg_fs_pred, car_test$acceptability)
```

---

# Part 2b: Improving Model Performance via Hyperparameter Tuning

## Tuning the Neural Network (`decay`)

```{r}
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
nn_grid <- expand.grid(size = 5, decay = c(0, 0.001, 0.005, 0.01, 0.02, 0.05, 0.1))
set.seed(123)

nn_tuned <- train(car_train[, -7], car_train[, 7], method = "nnet",
                   trControl = ctrl, tuneGrid = nn_grid, trace = FALSE)
plot(nn_tuned)
```

## Tuning the Decision Tree (`trials`)

```{r}
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
tree_grid <- expand.grid(model = "tree",
                          trials = c(1, 5, 10, 15, 20, 25, 30, 40, 50, 60, 80, 100),
                          winnow = FALSE)
set.seed(123)

tree_tuned <- train(car_train[, -7], car_train[, 7], method = "C5.0",
                     trControl = ctrl, tuneGrid = tree_grid)
plot(tree_tuned)
```

---

# Part 3. Presentation of Specified Results

*(This section will be finalised during the lab session once the specific 
figures to include are confirmed. Draft notes and interpretations for each 
required figure should be added here, along with Harvard-style references.)*

## References (Harvard style - to be completed/expanded)

- Kuhn, M. (2008) 'Building predictive models in R using the caret package', 
  *Journal of Statistical Software*, 28(5), pp. 1-26.
- Meyer, D. et al. (2023) *e1071: Misc functions of the Department of 
  Statistics, Probability Theory Group*. R package.
- Meyer, P. E. (2014) *infotheo: Information-theoretic measures*. R package.
- UCI Machine Learning Repository (1997) *Car Evaluation Data Set*. Available 
  at: https://archive.ics.uci.edu/dataset/19/car+evaluation (Accessed: 
  [add date]).
- Wickham, H. (2016) *ggplot2: Elegant Graphics for Data Analysis*. New York: 
  Springer-Verlag.
